Windowed least-squares model reduction for dynamical systems

### 2019-10-17 cavity.tex
latest

The reduced-order models leverage POD to construct the \spatialAcronym\ trial basis and use q-sampling~\cite{qdeim_drmac} based on snapshots of the FOM velocity to select the sampling points and, as a result the weighting 
matrix $\stweightingMatArg{}$; we use a constant weighting matrix across all windows, e.g., $\stweightingMatArg{n} \equiv \stweightingMatArg{}$. The process used to construct the initial conditions and trial subspaces for the ROMs is as follows:
\begin{enumerate}
\item Initialize the FOM with uniform free-stream conditions.
\item Evolve the FOM for $t \in [0,400]$.
\item Reset the time-coordinate, $0 \leftarrow t$, and execute Algorithm~\ref{alg:pod} with $\stateInterceptArg{} = \bz, N_{\text{skip}} = 100$ and $K = \{136,193\}$ over $t \in [0,100]$ to construct two trial subspaces comprising $\approx 95\%$ and $\approx 97\%$ of the snapshot energy, respectively. 
\item Execute Algorithm~\ref{alg:qdeim} with $N_{\text{skip}} = 100$, $n_s = 129$ to obtain the sampling point matrix of dimension $\stweightingMatOneArg{} \in \{0,1\}^{1603 \times \fomdim }$. 
\end{enumerate}
Figure~\ref{fig:cav_sampmesh} shows the resulting sample mesh used in the ROM simulations. To depict the nature of the flow, Figure~\ref{fig:fom_sols_cav} shows snapshots of the vorticity field generated by the FOM for several time instances used in training.  
\begin{table}[]
\begin{centering}
\begin{tabular}{c c c c}
\hline
Basis \# & Trial Basis Dimension ($K$) &  Energy Criterion & Sample Points ($n_S$) \\
\hline
%1    & $136$ & $95.0\%$ & $580$ \\
%2    & $193$ & $97.0\%$ & $580$ \\
1    & $136$ & $95.0\%$ & $1603$ \\
2    & $193$ & $97.0\%$ & $1603$ \\
\hline
\end{tabular}
\caption{Summary of the various basis sizes employed in the cavity flow example.}
\label{tab:rom_basis_details}
\end{centering}
\end{table}

\subsubsection{Description of reduced-order models}
Hyper-reduced ROMs based on the Galerkin, collocated Least--Squares Petrov--Galerkin, and \methodAcronym\ approaches are 
again considered. Details on the implementation of the methods is as follows:
\begin{itemize}
\item \textit{Galerkin ROM with collocation}: We consider a Galerkin ROM with collocation. Unless 
otherwise noted, the Galerkin ROM is evolved in time with the Crank--Nicolson time scheme at a time step of $\Delta t =0.1$.

\item \textit{LSPG ROM with collocation}: We consider a collocated LSPG ROM, which is built on top of the FOM discretization using the Crank--Nicolson scheme for temporal 
discretization. Unless otherwise noted, the time step for the LSPG ROM is $\Delta t  = 0.1$. As in the previous example, 
the Gauss--Newton method is used to solve the non-linear least-squares problem arising at each time instance. We deem the Gauss--Newton 
algorithm converged when the gradient norm is less than $1e-4$. All Jacobians are computed via automatic differentiation.

\item \textit{\methodAcronymROMs\ with collocation}: We consider \methodAcronymROMs\ solved via the direct method. The ROMs use the Crank--Nicolson time-discretization with a time step of 
$\Delta t = 0.1$. We deem the Gauss--Newton algorithm converged when the gradient norm is less than $1e-4$. All Jacobians are computed via automatic differentiation. Uniform quadrature 
weights are used for evaluating the integral in the minimization problem~\eqref{eq:obj_gen_slab}.

>>>>

The reduced-order models leverage POD to construct the \spatialAcronym\ trial basis and use q-sampling~\cite{qdeim_drmac} based on snapshots of the FOM velocity to select the sampling points (and as a result the weighting 
matrix $\stweightingMatArg{}$); we use a constant weighting matrix across all windows, e.g., $\stweightingMatArg{n} \equiv \stweightingMatArg{}$. The process used to construct the initial conditions and trial subspaces for the ROMs is as follows:
\begin{enumerate}
\item Initialize the FOM with uniform free-stream conditions.
\item Evolve the FOM for $t \in [0,400]$.
\item Reset the time coordinate, $0 \leftarrow t$, and execute Algorithm~\ref{alg:pod} with $\stateInterceptArg{} = \bz, N_{\text{skip}} = 100$ and $K = \{136,193\}$ over $t \in [0,100]$ to construct two trial subspaces comprising $\approx 95\%$ and $\approx 97\%$ of the snapshot energy, respectively. 
\item Execute Algorithm~\ref{alg:qdeim} with $N_{\text{skip}} = 100$, $n_s = 129$ to obtain the sampling point matrix of dimension $\stweightingMatOneArg{} \in \{0,1\}^{1603 \times \fomdim }$. 
\end{enumerate}
Figure~\ref{fig:cav_sampmesh} shows the resulting sample mesh used in the ROM simulations. To depict the nature of the flow, Figure~\ref{fig:fom_sols_cav} shows snapshots of the vorticity field generated by the FOM for several time instances used in training.  
\begin{table}[]
\begin{centering}
\begin{tabular}{c c c c}
\hline
Basis \# & Trial Basis Dimension ($K$) &  Energy Criterion & Sample Points ($n_S$) \\
\hline
%1    & $136$ & $95.0\%$ & $580$ \\
%2    & $193$ & $97.0\%$ & $580$ \\
1    & $136$ & $95.0\%$ & $1603$ \\
2    & $193$ & $97.0\%$ & $1603$ \\
\hline
\end{tabular}
\caption{Summary of the various basis sizes employed in the cavity flow example.}
\label{tab:rom_basis_details}
\end{centering}
\end{table}

\subsubsection{Description of reduced-order models}
We consider collocated ROMs based on the Galerkin, least-squares Petrov--Galerkin, and \methodAcronym\ approaches. Details on the implementation of the methods is as follows:
\begin{itemize}
\item \textit{Galerkin ROM with collocation}: We consider a Galerkin ROM with collocation. 
The Galerkin ROM is evolved in time with the CN time scheme at a time step of $\Delta t =0.1$.

\item \textit{LSPG ROM with collocation}: We consider a collocated LSPG ROM, which is built on top of the FOM discretization using the Crank--Nicolson scheme for temporal 
discretization. Unless otherwise noted, the time step for the LSPG ROM is $\Delta t  = 0.1$. As in the previous example, 
the Gauss--Newton method is used to solve the nonlinear least-squares problem arising at each time instance. The linear least-squares problems 
arising at each Gauss--Newton iteration are solved via the normal equations. We deem the Gauss--Newton 
algorithm converged when the gradient norm is less than $1e-4$. All Jacobians are computed via automatic differentiation.

\item \textit{\methodAcronymROMs\ with collocation}: We consider \methodAcronymROMs\ solved via the direct method. The ROMs use the Crank--Nicolson time-discretization with a time step of 
$\Delta t = 0.1$. We deem the Gauss--Newton algorithm converged when the gradient norm is less than $1e-4$. The linear least-squares problems
arising at each Gauss--Newton iteration are solved via the normal equations. All Jacobians are computed via automatic differentiation. Uniform quadrature 
weights are used for evaluating the integral in the minimization problem~\eqref{eq:obj_gen_slab}.

### 2019-10-17 numerical_experiments.tex
latest

\subsubsection{Numerical results}
The impact of the window size on \methodAcronymROMs\ is first assessed. We consider a set of \methodAcronymROMs\ that minimize the residual over windows of constant size 
$\DeltaSlabArg{n} \equiv \DeltaSlabArg{} =  .002$, $0.004$ ,$0.008$, $0.02$, $0.04$, $0.10$, $0.20$, $1.0$. We additionally consider the standard Galerkin and LSPG ROMs. We first show results for \methodAcronymROMs\ using the direct method with CN time discretization; a comparison of different time-marching methods and direct/indirect solution techniques will be provided later in this section. 
First, Figure~\ref{fig:sod_density} presents the density solutions predicted by the various ROMs at $t = 0.5$ and $1.0$. Figure~\ref{fig:sod_xt} shows 
$x-t$ diagrams for the same density solutions. From Figures~\ref{fig:sod_density} and~\ref{fig:sod_xt}, it is observed that the LSPG and Galerkin ROMs accurately characterize 
the system: they correctly track the shock location, expansion waves, etc. We observe both predictions, however, to be highly oscillatory. These oscillations are 
not physical and can lead to numerical blowup; e.g., due to negative pressure. We observe the \methodAcronymROMs\ to produce less oscillatory solutions than both 
the Galerkin and LSPG ROMs. Critically, we see that the solution becomes less oscillatory as the window size over which the residual is minimized grows. The solution displays no oscillations when the residual is minimized over the entire space--time domain.

@@

Next, Figure~\ref{fig:sod_walltime} provides the CPU wall-clock times for the various \methodAcronymROMs. The computational cost of all methods is seen to grow 
as the window size is increased. For indirect methods, this increase in cost is due to the fact that, as the window size 
grows, more iterations of the FBSM are required for convergence. For direct methods, the increase in cost is due to (1) the cost 
associated with forming and solving the normal equations at each Gauss--Newton iteration and (2) the increased number of 
Gauss--Newton iterations required for convergence. The cost of the FBSM method is observed to increase more rapidly 
than the direct methods. Encouragingly, \methodAcronymROMs\ based on the CN discretization
minimizing the full space--time residual (comprising 
500 time instances) only  
cost approximately one order of magnitude more than the case where the residual is minimized over a single time-step (i.e., LSPG). Also of interest is the 
fact that the direct method utilizing the AB2 time-discretization scheme costs only slightly more per a given window size than the direct method 
utilizing the CN time discretization. This is despite the fact that AB2 is evolved at a time step of $\Delta t = 0.0005$, while CN uses $\Delta t = 0.002$; thus 
AB2 contains 4 times more temporal degrees of freedom than CN.  Finally, we emphasize that the 
results presented here are presented for standard algorithms (e.g., Guass--Newton and the FBSM). As mentioned in 
Remarks~\ref{remark:gaussnewton} and~\ref{remark:fbsm}, we expect the computational cost 
of both the indirect and direct methods can be decreased through the use and/or development of 
algorithms tailored to the minimization problem.

@@

Finally, we study the impact of the time step on the \methodAcronymROM\ results. We examine \methodAcronymROMs\ that use a
window size of $\Delta T = 0.1$, with time steps $\Delta t =$  $0.001$, $0.002$, $0.005$, $0.01$, $0.05$, $0.1$ (i.e.,  
$100$, $50$, $20$, $10$, $2$, and $1$ time-steps per window). We additionally 
consider LSPG ROMs leveraging the same set of time steps. To assess time step convergence, we compare results to a new full-order model, which is as described in Section~\ref{sec:sod_fom} but uses a fine time step of $\Delta t = 1e-4$. Figure~\ref{fig:time_step_study} shows the space--time $\elltwo$ error and residual norm  
of the various ROMs. We observe that both the space--time $\elltwo$ error and residual norm 
of the \methodAcronymROMs\ decrease and converge as the time step decreases. This is in direct contrast to the LSPG ROMs, in where both the $\elltwo$ 
space--time error and residual norm increase as the time step decreases. This result 
demonstrates that \methodAcronym\ overcomes the time-step dependence inherit to the LSPG approach. Lastly, Figure~\ref{fig:walltime_dtvar}
shows the relative wall-clock times of the \methodAcronymROMs\ with respect to the LSPG ROMs. It is seen that, for all time steps considered, the \methodAcronym-ROMs 
are less than 6x the cost LSPG; despite the window sizes comprising up to 100 time steps. 
\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_vs_window_dtvar.pdf}
\caption{Space--time $\elltwo$ error of the \methodAcronymROM.}
\label{fig:sod_error_methods_a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/objective_vs_window_dtvar.pdf}
\caption{Space--time residual norm of the \methodAcronymROM.} 
\label{fig:sod_error_methods_b}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs.} 
\label{fig:time_step_study}
\end{center}
\end{figure}

@@

\subsubsection{Summary of numerical results for the Sod shock tube}
The key observations from the results of the first numerical example are: 
\begin{enumerate}
\item Increasing the window size over which the residual is minimized led to more physically relevant solutions. Specifically, we observed that as the window size over which the residual was minimized grew, \methodAcronym\ led to less oscillatory solutions.
\item Increasing the window size over which the residual is minimized does not necessary lead to a lower space--time state error as measured by the $\elltwo$ norm. We observed that minimizing the residual over an intermediary window size led to the lowest space--time state error in the $\elltwo$ norm. 
\item The \methodAcronym\ displays time-step convergence: both the space--time $\elltwo$ state error and residual norm decreased and displayed time-step convergence as the time step decreased. This is in contrast to LSPG.
\item For all examples considered, in where the windows comprised up to 2000 time instances, \methodAcronym\ with the direct method is between 1x and 10x the cost of the LSPG. The 
direct method was observed to be, in general, slightly more efficient and robust than the indirect method.
\end{enumerate}

>>>>

\subsubsection{Numerical results}
We first assess the impact of the window size on the performance of the \methodAcronymROMs.  We consider a set of \methodAcronymROMs\ that minimize the residual over windows of constant size 
$\DeltaSlabArg{n} \equiv \DeltaSlabArg{} =  .002$, $0.004$ ,$0.008$, $0.02$, $0.04$, $0.10$, $0.20$, $1.0$. We additionally consider the standard Galerkin and LSPG ROMs. We first show results for \methodAcronymROMs\ using the direct method with CN time discretization; a comparison of different time-marching methods and direct/indirect solution techniques will be provided later in this section. 
First, Figure~\ref{fig:sod_density} presents the density solutions predicted by the various ROMs at $t = 0.5$ and $1.0$. Figure~\ref{fig:sod_xt} shows 
$x-t$ diagrams for the same density solutions. From Figures~\ref{fig:sod_density} and~\ref{fig:sod_xt}, we observe that the LSPG and Galerkin ROMs accurately characterize 
the system: they correctly track the shock location, expansion waves, etc. We observe both predictions, however, to be highly oscillatory. These oscillations are 
not physical and can lead to numerical blowup; e.g., due to negative pressure. We observe the \methodAcronymROMs\ to produce less oscillatory solutions than both 
the Galerkin and LSPG ROMs. Critically, we see that the solution becomes less oscillatory as the window size over which the residual is minimized grows. The solution displays no oscillations when the residual is minimized over the entire space--time domain.

@@

Next, Figure~\ref{fig:sod_walltime} provides the CPU wall-clock times for the various \methodAcronymROMs. The computational cost of all methods is seen to grow 
as the window size is increased. For indirect methods, this increase in cost is due to the fact that, as the window size 
grows, more iterations of the FBSM are required for convergence. For direct methods, the increase in cost is due to (1) the cost 
associated with forming and solving the normal equations at each Gauss--Newton iteration and (2) the increased number of 
Gauss--Newton iterations required for convergence. The cost of the FBSM method is observed to increase more rapidly 
than the direct methods. Encouragingly, \methodAcronymROMs\ based on the CN discretization
minimizing the full space--time residual (comprising 
500 time instances) only  
cost approximately one order of magnitude more than the case where the residual is minimized over a single time step (i.e., LSPG). Also of interest is the 
fact that the direct method utilizing the AB2 time-discretization scheme costs only slightly more per a given window size than the direct method 
utilizing the CN time discretization. This is despite the fact that AB2 is evolved at a time step of $\Delta t = 0.0005$, while CN uses $\Delta t = 0.002$; thus 
AB2 contains 4 times more temporal degrees of freedom than CN.  Finally, we emphasize that the 
results presented here are presented for standard algorithms (e.g., Guass--Newton and the FBSM). As mentioned in 
Remarks~\ref{remark:gaussnewton} and~\ref{remark:fbsm}, we expect the computational cost 
of both the indirect and direct methods can be decreased through the use and/or development of 
algorithms tailored to the windowed minimization problem.

@@

Finally, we study the impact of the time step on the \methodAcronymROM\ results. We examine \methodAcronymROMs\ that use a
window size of $\Delta T = 0.1$, with time steps $\Delta t =$  $0.001$, $0.002$, $0.005$, $0.01$, $0.05$, $0.1$ (i.e.,  
$100$, $50$, $20$, $10$, $2$, and $1$ time instances per window). We additionally 
consider LSPG ROMs leveraging the same set of time steps. To assess time step convergence, we compare results to a new full-order model, which is as described in Section~\ref{sec:sod_fom} but uses a fine time step of $\Delta t = 10^{-4}$. Figure~\ref{fig:time_step_study} shows the space--time $\elltwo$ error and residual norm  
of the various ROMs. We observe that both the space--time $\elltwo$ error and residual norm 
of the \methodAcronymROMs\ decrease and converge as the time step decreases. This is in direct contrast to the LSPG ROMs, in where both the $\elltwo$ 
space--time error and residual norm increase as the time step decreases. This result 
demonstrates that \methodAcronym\ overcomes the time-step dependence inherit to the LSPG approach. Lastly, Figure~\ref{fig:walltime_dtvar}
shows the relative wall-clock times of the \methodAcronymROMs\ with respect to the LSPG ROMs. It is seen that for all time steps considered, the \methodAcronymROMs\ 
are less than 6x the cost of LSPG, despite the window sizes comprising up to 100 time instances. 
\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_vs_window_dtvar.pdf}
\caption{Space--time $\elltwo$ error of the \methodAcronymROM.}
\label{fig:sod_error_methods_a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/objective_vs_window_dtvar.pdf}
\caption{Space--time residual norm of the \methodAcronymROM.} 
\label{fig:sod_error_methods_b}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs.} 
\label{fig:time_step_study}
\end{center}
\end{figure}

@@

\subsubsection{Summary of numerical results for the Sod shock tube}
The key observations from the results of the first numerical example are: 
\begin{enumerate}
\item Increasing the window size over which the residual is minimized led to more physically relevant solutions. Specifically, we observed that as the window size over which the residual was minimized grew, \methodAcronym\ led to less oscillatory solutions.
\item Increasing the window size over which the residual is minimized does not necessary lead to a lower space--time state error as measured by the $\elltwo$ norm. We observed that minimizing the residual over an intermediary window size led to the lowest space--time state error in the $\elltwo$ norm. 
\item \methodAcronym\ displays time-step convergence: both the space--time $\elltwo$ state error and residual norm decreased and displayed time-step convergence as the time step decreased. This is in contrast to LSPG.
\item For all examples considered, in where the windows comprised up to 2000 time instances, \methodAcronym\ with the direct method is between 1x and 10x the cost of the LSPG. The 
direct method was observed to be, in general, slightly more efficient and robust than the indirect method.
\end{enumerate}
